<!--
  %\VignetteEngine{knitr}
  %\VignetteIndexEntry{hypothetical example}
  %\VignetteDepends{categoryComparePaperRev, ggplot2}
-->
  
# Hypothetical Example
  
To demonstrate the utility of the general `categoryCompare` approach, we will construct an artificial data set based on the Gene Ontology. We will attempt to demonstrate that differences in the **set** or **list** based comparisons are also dependent on the number of items annotated to a particular term.

## GO Terms

```{r nGO}
nGO <- 20
```

We will make use of three sets of `r nGO` GO terms from *H. sapiens*. For each set, we want differing numbers of genes annotated to them.

### Num Annotations Dist

Let's first look at the distribution of number of genes annotated to the GO terms in **biological process**

```{r goAnnotations}
library(org.Hs.eg.db)
library(GO.db)
hsGO <- as.list(org.Hs.egGO2ALLEGS)

goOntology <- Ontology(names(hsGO))
goBP <- names(goOntology)[goOntology == "BP"]
hsGO <- hsGO[goBP]
hsGO <- lapply(hsGO, unique)
```

```{r goCountDist}
library(ggplot2)
hsGO_count <- sapply(hsGO, length)
hsGO_count <- data.frame(count=hsGO_count)
ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
```

### Set up groups

```{r defineMin}
minGO <- 10
```

Let us define a minimum number of genes that need to be annotated to a GO term (`r minGO`), and then divide the GO terms into groups that we can subsequently take random samples from.

```{r defineGroupLimits}
grpLow <- c(10, 100)
grpMed <- c(250, 500)
grpHi <- c(500, 1500)
```

```{r sampleGroups}
set.seed(271113) # for reproducibility
nGO <- 20

GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
```

