<!--
  %\VignetteEngine{knitr}
  %\VignetteIndexEntry{hypothetical example}
  %\VignetteDepends{categoryComparePaperRev, ggplot2}
-->
  
# Hypothetical Example
  
To demonstrate the utility of the general `categoryCompare` approach, we will construct an artificial data set based on the Gene Ontology. We will attempt to demonstrate that differences in the **set** or **list** based comparisons are also dependent on the number of items annotated to a particular term.

## GO Terms

```{r nGO}
nGO <- 20
```

We will make use of three sets of `r nGO` GO terms from *H. sapiens*. For each set, we want differing numbers of genes annotated to them.

### Num Annotations Dist

Let's first look at the distribution of number of genes annotated to the GO terms in **biological process**

```{r goAnnotations}
library(org.Hs.eg.db)
library(GO.db)
library(ccPaperRev)
hsGO <- as.list(org.Hs.egGO2ALLEGS)

goOntology <- Ontology(names(hsGO))
goBP <- names(goOntology)[goOntology == "BP"]
hsGO <- hsGO[goBP]
hsGO <- lapply(hsGO, unique)
```

```{r goCountDist}
library(ggplot2)
hsGO_count <- sapply(hsGO, length)
hsGO_count <- data.frame(count=hsGO_count)
ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
```

### Set up groups

```{r defineMin}
minGO <- 10
```

Let us define a minimum number of genes that need to be annotated to a GO term (`r minGO`), and then divide the GO terms into groups that we can subsequently take random samples from.

```{r defineGroupLimits}
grpLow <- c(10, 100)
grpMed <- c(250, 500)
grpHi <- c(500, 1500)
```

```{r sampleGroups}
set.seed(271113) # for reproducibility
nGO <- 20

GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
```

### Get Genes

Now, for all of these we need to generate **two** samples of genes from them to comprise our **differentially expressed** (DE) sets from the genome, representing two different expression experiments.

How do we select these two sets of genes?

* define the maximum number of diff genes for the set
* Randomly sort the GO terms
* For each GO term in turn (i.e. iterating over the GO terms):
  * take a random fraction of genes for that GO term (sampling from a log-normal distribution of possibilites), adding the requisite number of genes
  * If all the genes for a GO term get taken, remove that GO term from future consideration
* Continue iterating through until the total number of genes is reached

What we expect from this is that in both datasets, the GO terms with lower numbers of genes annotated will have large fractions of their annotations present, while GO terms with more and more genes will have lower fractions, and there will be less overlap between the two datasets.

These expectations should be checked.

```{r generateSamples}
useGO <- c(GO_low, GO_med, GO_hi)
sample1 <- sampleTerms(useGO, hsGO, 2000, 4)
sample2 <- sampleTerms(useGO, hsGO, 2000, 4)
```

```{r checkAssumptions}
# check assumptions
useGO_counts <- sapply(hsGO[useGO], length)
useGO_frac <- sapply(hsGO[useGO], function(x){
  length(intersect(x, sample1)) / length(x)
})
checkGO <- data.frame(counts=useGO_counts, frac=useGO_frac, sample="1")
useGO_frac2 <- sapply(hsGO[useGO], function(x){
  length(intersect(x, sample2)) / length(x)
})
checkGO <- rbind(checkGO, data.frame(counts=useGO_counts, frac=useGO_frac2, sample="2"))
ggplot(checkGO, aes(x=counts, y=frac, color=sample)) + geom_point()
```

From the plot above, it appears that we meet expectations. In addition, we manage to share only `r length(intersect(sample1, sample2))` genes between the two samples. 

What about the degree of overlap between the two samples as a function of size or fraction?

```{r overlapDist}
overlapDist <- sapply(useGO, function(x){
  baseAnn <- hsGO[[x]]
  t1 <- intersect(baseAnn, sample1)
  t2 <- intersect(baseAnn, sample2)
  c(length(baseAnn), length(t1), length(t2), length(intersect(t1,t2)))
})
overlapDist <- t(overlapDist)
overlapFrac <- c(overlapDist[,2] / overlapDist[,1], overlapDist[,3]/overlapDist[,1], overlapDist[,4]/overlapDist[,1])
overlapFrac <- data.frame(count=rep(overlapDist[,1], 3), 
                          frac=overlapFrac, 
                          id=c(rep("s1", 60), rep("s2", 60), rep("s1.s2", 60)))
overlapFrac
ggplot(overlapFrac, aes(x=count, y=frac, color=id)) + geom_point()
```

