<!--
  %\VignetteEngine{knitr}
  %\VignetteIndexEntry{hypothetical example}
  %\VignetteDepends{categoryComparePaperRev, ggplot2}
-->
  
# Hypothetical Example
  
To demonstrate the utility of the general `categoryCompare` approach, we will construct an artificial data set based on the Gene Ontology. We will attempt to demonstrate that differences in the **set** or **list** based comparisons are also dependent on the number of items annotated to a particular term.

## Definitions

**Set** based: calculations performed independently on each *set* of features, and then the results are combined.

**List** based: calculations performed on the intersected *list* of two sets of features.

## GO Terms

```{r nGO}
nGO <- 20
```

We will make use of three sets of `r nGO` GO terms from *H. sapiens*. For each set, we want differing numbers of genes annotated to them.

### Num Annotations Dist

Let's first look at the distribution of number of genes annotated to the GO terms in **biological process**

```{r goAnnotations}
library(org.Hs.eg.db)
library(GO.db)
library(ccPaperRev)
hsGO <- as.list(org.Hs.egGO2ALLEGS)

goOntology <- Ontology(names(hsGO))
goBP <- names(goOntology)[goOntology == "BP"]
hsGO <- hsGO[goBP]
hsGO <- lapply(hsGO, unique)
universeGenes <- unique(unlist(hsGO))
```

```{r goCountDist}
library(ggplot2)
hsGO_count <- sapply(hsGO, length)
hsGO_count <- data.frame(count=hsGO_count)
ggplot(hsGO_count, aes(x=count)) + geom_bar(binwidth=10) + xlim(0, 2000) + ylim(0, 500)
```

### Set up groups

```{r defineMin}
minGO <- 10
```

Let us define a minimum number of genes that need to be annotated to a GO term (`r minGO`), and then divide the GO terms into groups that we can subsequently take random samples from.

```{r defineGroupLimits}
grpLow <- c(10, 100)
grpMed <- c(250, 500)
grpHi <- c(500, 1500)
```

```{r sampleGroups}
set.seed(271113) # for reproducibility
nGO <- 20

GO_low <- limitedRandomSample(hsGO_count, grpLow, nGO)
GO_med <- limitedRandomSample(hsGO_count, grpMed, nGO)
GO_hi <- limitedRandomSample(hsGO_count, grpHi, nGO)
```

### Get Genes

Now, for all of these we need to generate **two** samples of genes from them to comprise our **differentially expressed** (DE) sets from the genome, representing two different expression experiments.

How do we select these two sets of genes?

* define the maximum number of diff genes for the set
* Randomly sort the GO terms
* For each GO term in turn (i.e. iterating over the GO terms):
  * take a random fraction of genes for that GO term (sampling from an exponentially decaying distribution of possibilites), adding the requisite number of genes
  * If all the genes for a GO term get taken, remove that GO term from future consideration
* Continue iterating through until the total number of genes is reached

What we expect from this is that in both datasets, the GO terms with lower numbers of genes annotated will have large fractions of their annotations present, while GO terms with more and more genes will have lower fractions, and there will be less overlap between the two datasets.

These expectations should be checked.



```{r generateSamples}
useGO <- c(GO_low, GO_med, GO_hi)
goList <- list(low=GO_low, med=GO_med, hi=GO_hi)
nGene <- 1000
sample1_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
sample2_org <- sampleTerms(useGO, hsGO, 1000, 4)[1:nGene]
```

### Checking our expectations about the fractions of GO terms.

```{r checkFractions}
goFractions <- calcFraction(hsGO[useGO], list(sample1=sample1_org, sample2=sample2_org))
ggplot(goFractions, aes(x=size, y=frac, color=genelist)) + geom_point()
```

### Do calculations

For each `geneList`, we will do the GO enrichment calculations, and then get the results. We also do the intersection of all the `geneList`'s, and do the calculations again.

```{r runCalcs}
samples_noNoise <- list(sample1 = sample1_org, sample2 = sample2_org)
go_noNoise <- hyperGOMultiEnrichment(samples_noNoise, universe=universeGenes)
```

Now get the p-values out (`-1 * log10(pvalue)`). For each of our test GO terms, we take the minimum p-value from the results of the **set** calculations, and also the values from the **list** calculations.

```{r noNoisePvalues}
noNoise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noNoise)

noNoise_pvalues <- pvalueDiffSig(noNoise_pvalues, pCutoff=0.05, log=TRUE)
```

Finally, lets add in our values of the fraction, total genes, and the class (**low**, **med**, **hi**) so we can do some fancy plotting stuff with it.

```{r noNoiseAddInfo}
sizeClass <- rep(c("low", "med", "hi"), each=20)
names(sizeClass) <- useGO
noNoise_pvalues$sizeClass <- sizeClass
noNoise_pvalues$size <- goFractions[1:60,4]
noNoise_pvalues$frac <- goFractions$frac[1:60]
```

Now we will plot the difference in log-p-values (where we have **set** - **list**), so that positive values imply that **set** had lower p-values than **list**, based on the *fraction* of genes annotated.

```{r noNoisePlotStuff}
noNoise_pvalues$sizeClass <- factor(noNoise_pvalues$sizeClass, levels=c("low", "med", "hi"), ordered=TRUE)
ggplot(noNoise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
ggplot(noNoise_pvalues, aes(x=sizeClass, y=diff)) + geom_boxplot() + geom_point()
```

We can see that it appears in general that **set** has better p-values than **list**, especially in the *med* and *hi* groups. Also, both methods get pretty much the same GO terms as significant. 

How many of the GO terms have better p-values in **set** compared to **list**?

```{r noNoiseBetterPVals}
tapply(noNoise_pvalues$diff, noNoise_pvalues$sizeClass, function(x){sum(x > 0)})
```

What is the average difference among each group?

```{r noNoiseBetterMean}
tapply(noNoise_pvalues$diff, noNoise_pvalues$sizeClass, mean)
```

So ideally, the mean of the full set of differences, combined with how many pass in each case, should adequately quantify how well each of the methods is doing. 

## Noise

In addition to the genes that are annotated to the GO terms of interest, we will also add in genes that have no relation to the GO terms under consideration, these would be considered **noise** genes. To make the model simple, we will have the **same** genes as noise in **both** samples.

```{r noiseGenes}
nNoise <- 500
gene2HsGO <- reverseSplit(hsGO)
not_useGO <- sapply(gene2HsGO, function(x){
  sum(x %in% useGO) == 0
})
noiseGenes <- names(not_useGO)[not_useGO]
noiseGenes <- sample(noiseGenes, nNoise)

# check that we did this right, the fraction should not change after adding noise genes
sample1 <- c(sample1_org, noiseGenes)
sample2 <- c(sample2_org, noiseGenes)

samplesNoise <- list(sample1=sample1, sample2=sample2)

goFracNoise <- calcFraction(hsGO[useGO], samplesNoise)
plot(goFracNoise$frac, goFractions$frac)
```


```{r noisyEnrichment}
go_noise <- hyperGOMultiEnrichment(samplesNoise, universeGenes)
```

```{r noisePvalues}
noise_pvalues <- pvaluesMultiEnrich(c("sample1", "sample2"), useGO, go_noise)

noise_pvalues <- pvalueDiffSig(noise_pvalues, pCutoff=0.05, log=TRUE)

noise_pvalues$sizeClass <- sizeClass
noise_pvalues$size <- goFracNoise[1:60,4]
noise_pvalues$frac <- goFracNoise$frac[1:60]
```

```{r plotSummarizeNoise}
noise_pvalues$sizeClass <- factor(noise_pvalues$sizeClass, c("low", "med", "hi"), ordered=TRUE)
ggplot(noise_pvalues, aes(x=frac, y=diff, color=sigState)) + geom_point() + facet_grid(. ~ sizeClass, scales="free_x")
```

```{r summarizeCountsNoise}
tapply(noise_pvalues$diff, noise_pvalues$sizeClass, function(x){sum(x > 0)})
tapply(noise_pvalues$diff, noise_pvalues$sizeClass, mean)
```

Although using gene lists that were perfect (i.e. no noise) everything in the **list** method was pretty much still significant, on average **sets** gave much better p-values. However, as we add noise to the system in the form of genes that are not annotated to our test GO terms, more terms are significant only in the **sets** method (i.e. show up in **both** sets). 


## Expectations

Based on the above results, what I expect we will see if we do a programmatic search over the space is that there is a dependence on both the fraction of DE genes that are noisy, and the fraction of noisy genes that are shared (which really amounts to the same thing, but is a subtle point). 