# Crohn's VS Ulcerative Colitis

GEO dataset **GSE36807** has crohn's, uc, and normal samples. 

```{r getData, eval=FALSE}
library(GEOquery)

cucData <- getGEO(GEO="GSE36807")[[1]]

tmpP <- pData(cucData)
tmpP$status <- "control"

cLoc <- grep("Crohn", tmpP$characteristics_ch1)
tmpP$status[cLoc] <- "crohns"

ucLoc <- grep("Ulcerative colitis", tmpP$characteristics_ch1)
tmpP$status[ucLoc] <- "uc"
pData(cucData) <- tmpP

.sessionInfo <- sessionInfo()
.timeDate <- Sys.time()
save(cucData, .sessionInfo, .timeDate, file="inst/data/uc_crohns_rawData.RData")
```

## Differential analysis

Use the `limma` package to rank probes. 

```{r rankProbes}
library(limma)
library(hgu133plus2.db)
load("inst/data/uc_crohns_rawData.RData")

cucComps <- c("uc - control", "crohns - control")
geneID <- unlist(mget(featureNames(cucData), hgu133plus2ENTREZID))

cucExpr <- exprs(cucData)
cucCharacter <- pData(cucData)

cucFC <- rankGenes(cucExpr, cucCharacter$status, cucComps, dupStrategy="minP", aggregateIndex=geneID)
names(cucFC) <- c("UC", "CROHNS")

cucDiff <- lapply(cucFC, getDiffGenes)
# this object can be run using base ccEnrich, or can be easily coerced
```


## Run GSEA type analysis using limma

There are two ways to do GSEA using `limma`. 

  1 Use `romer` for competitive tests with rotations to do random samples
  2 Use `geneSetTest` for gene-wise competitive tests, *i.e.* each set is tested against random gene samples
  
We're going to try both and see what we get out.

```{r runGSEA}
library(GO.db)
hsGO <- as.list(org.Hs.egGO2ALLEGS)
hsGO <- hsGO[(Ontology(names(hsGO)) == "BP")]


```

